Acceleration of Global Search by Implementing Dual Estimates for Lipschitz Constant\thanks{This research was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2019-09-10 2019/NUMTA_Barkalov/LNCS/numta19.tex
Semi-final

The assumption of the objective function to be Lipschitzian is typical of many approaches to the development of the global optimization algorithms \cite{Evtushenko2013,Zilinskas2010,Pinter1996,Strongin2000}. Moreover, the adaptive estimate of the unknown Lipschitz constant, based on the obtained search information, is one of the most important problems being solved in these algorithms. 
The value of the Lipschitz constant affects essentially the convergence rate of the global optimization algorithms. Therefore, the issue of its correct estimate is so important. 
The underestimation of the real value of this constant may result in losing the convergence of the algorithm to the global solution. At the same time, if the value of the constant estimate is too large, the objective function may not match its real behavior. This results in  slow convergence of the algorithm to the global minimizer.

@@

In this paper, we follow the approach based on the idea of reducing the dimension with the use of the Peano-Hilbert curves \cite{Sergeyev2013,Strongin2000}, which continuously and unambiguously map the unit interval $[0,1]$ onto the $N$-dimensional cube $D$ from (\ref{D}) has been used. By using this kind of mapping, it is possible to reduce the multidimensional problem (\ref{problem}) to a univariate problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}},
\]
where the function $\varphi(y(x))$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
with the H{\"o}lder constant $H$ linked to the Lipschitz constant $L$ by the relation
$ H=2 L \sqrt{N+3}$ and $y(x)$ is a Peano-Hilbert curve from $[0,1]$ onto $D$.
%Russian
Note that theoretically the Peano-Hilbert curve $y(x)$ определяется as a limit object. is defined as a limit object. Therefore, in practical implementation, only some approximation to the true space-filling curve can be constructed. Some methods for constructing this type of approximations (called evolvents) are considered in \cite{Sergeyev2013,Strongin2000}. При этом точность приближения развертки к истинной кривой $y(x)$ зависит от плотности развертки $m$ (являющейся параметром ее построения) и составляет величину порядка $2^{-m}$ по каждой координате.

@@

The global search algorithm presented in the previous section is intended for solving the multiextremal problems, in which the objective function satisfies the Lipschitz condition. It is not necessary to define the value of the constant for the algorithm convergence. The estimation of the constant is performed in the course of global search based on available search information. 
According to the theorem from \cite{Strongin2000}, the sequence of the trial points $\{y^k\}$ will converge to the global minimizer $y^*$ if the condition 
\begin{equation}\label{cond}
r\mu > 2^{3-1/N}L\sqrt{N+3}.
\end{equation}
\noindent is satisfied. Thus,  an appropriate choice of the parameter $r$ from (\ref{R}) allows using the value $(r\mu) / (2^{3-1/N}\sqrt{N+3})$ as an estimate of the Lipschitz constant for the objective function $\varphi(y)$.

@@

An approach, in which two estimates of the Lipschitz constant %, one of which is essentially greater than another, 
are used in the rules of the algorithm, seems quite promising. 
This approach implies the use in the algorithm of two parameters $r_{glob}$ and $r_{loc}$ where $r_{glob} > r_{loc}>1$.

@@

This method for computing the characteristic is substantiated as follows. 
%Russian
Each search iteration will yield an interval with the current minimum value $z^*$ from (\ref{z}) at one of its boundaries. In the final phase of the search, this interval will correspond to the interval containing the global minimum, i.e. it will be the best one in terms of conducting further trials within it.

Let the current minimum value of $z^*$ from (\ref{z}) be achieved at the left point of the $i^{\rm th}$ interval, i.e. $z^* = z_{i-1}$. 
As proven in \cite{Strongin2000}, according to the rule (\ref{R}) the following inequality will be true: 
\[
R(i) \geq \Delta_i \left( 1 - \frac{1}{r} \right)^2.
\]
Therefore, for evaluating the characteristics $R_{loc}(i)$ and $R_{glob}(i)$ calculated with different parameters $r_{loc}$ and $r_{glob}$, the following relation will hold:
\[
	\Delta_i \left( 1 - \frac{1}{r_{glob}} \right)^2 > \Delta_i \left( 1 - \frac{1}{r_{loc}} \right)^2.
\]

Thus, when choosing the largest of the characteristics in accordance with
\[
R(i) = \max\{R_{loc}(i),R_{glob}(i)\}
\]
%as a rule,
the characteristic $R_{glob}(i)$ corresponding to the higher estimate of the Lipschitz constant will be chosen; the lower estimate (which speeds up the process of refining the current solution) will not be used.
However, if we multiply the lower estimate for the characteristic $R_{loc}(i)$ by the coefficient $\rho$ in accordance with (\ref{pho}),then such lower estimates will be equal, 
thus the choice of the characteristic $R_{loc}(i)$ corresponding to the lower estimate of the Lipschitz constant will become more likely.

@@

The advantages of the GSA-DL algorithm over its prototype are also confirmed by the operational characteristics of the algorithms as well.

Assume a series of test problems to be solved. The results of solving the series can be presented by a function $p(k)$ featuring the fraction of the total number of problems solved in $k$ iterations. Such a function will be called the \textit{operational characteristic} of the algorithm.

@@

The lower curves in Fig. \ref{oper4} and Fig. \ref{oper5} feature the characteristics of GSA whereas the upper ones, those of the GSA-DL. Such relative positions of the curves show the algorithm with two estimates of the Lipschitz constant is much faster on average when solving the problem series than the algorithm using a single estimate of the constant.
%Russian
Note that to solve problems of all classes (except for the \textit{Hard} class at $N=5$) the GSA-DL method requires about half as many trials as the GSA. The deterioration of the results in the case of the \textit{Hard} class at $N=5$ is explained by the complexity of this class' functions, which have a large attraction region of local minima and a narrow attraction region of the global minimum. To correctly solve such problems, the GSA-DL method often uses a higher estimate of the Lipschitz constant, thus reducing the speed difference of the GSA and GSA-DL algorithms.

>>>>

The assumption of the objective function to be Lipschitzian is typical of many approaches to the development of the global optimization algorithms \cite{Evtushenko2013,Zilinskas2010,Pinter1996,Strongin2000}. Moreover, the adaptive estimate of the unknown Lipschitz constant, based on the obtained search information, is one of the most important problems being solved in these algorithms. 
The value of the Lipschitz constant affects essentially the convergence rate of the global optimization algorithms. Therefore, the issue of its correct estimate is so important. 
The underestimation of the real value of this constant may result in losing the convergence of the algorithm to the global solution. At the same time, if the value of the constant estimate for the objective function is too large and does not match its real behavior, this will slow down the convergence of the algorithm to the global minimizer.

@@

In this paper, we follow the approach based on the idea of reducing the dimension with the use of the Peano-Hilbert curves \cite{Sergeyev2013,Strongin2000}, which continuously and unambiguously map the unit interval $[0,1]$ onto the $N$-dimensional cube $D$ from (\ref{D}). By using this kind of mapping, it is possible to reduce the multidimensional problem (\ref{problem}) to a univariate problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}},
\]
where the function $\varphi(y(x))$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N}
\]
with the H{\"o}lder constant $H$ linked to the Lipschitz constant $L$ by the relation
$ H=2 L \sqrt{N+3}$ and $y(x)$ is a Peano-Hilbert curve from $[0,1]$ onto $D$.
%Russian
Note that theoretically the Peano-Hilbert curve $y(x)$ is defined as a limit object. Therefore, in practical implementation, only some approximation to the true space-filling curve can be constructed. Some methods for constructing this type of approximations (called evolvents) are considered in \cite{Sergeyev2013,Strongin2000}. In this case, the accuracy of the evolvent approximation to the true curve $y(x)$ depends on the density of the evolvent $m$ (which is a parameter for constructing the evolvent) and is of the order of $2^{-m}$ for each coordinate.

@@

The global search algorithm presented in the previous section is intended for solving the multiextremal problems, in which the objective function satisfies the Lipschitz condition. It is not necessary to define the value of the constant for the algorithm convergence. The estimation of the constant is performed in the course of global search based on available search information. 
According to the theorem from \cite{Strongin2000}, the sequence of the trial points $\{y^k\}$ will converge to the global minimizer $y^*$ if the condition 
\begin{equation}\label{cond}
r\mu > 2^{3-1/N}L\sqrt{N+3}
\end{equation}
\noindent is satisfied. Thus,  an appropriate choice of the parameter $r$ from (\ref{R}) allows using the value $(r\mu) / (2^{3-1/N}\sqrt{N+3})$ as an estimate of the Lipschitz constant for the objective function $\varphi(y)$.

@@

An approach, in which two estimates of the Lipschitz constant %, one of which is essentially greater than another, 
are used in the rules of the algorithm, seems quite promising. 
This approach implies the use in the algorithm of two parameters $r_{glob}$ and $r_{loc}$, where $r_{glob} > r_{loc}>1$.

@@

This method for computing the characteristic can be substantiated as follows. 
%Russian
Each search iteration will yield an interval with the current minimum value $z^*$ from (\ref{z}) at one of its boundaries. In the final phase of the search, this interval will correspond to the interval containing the global minimum, i.e. it will be the best one in terms of conducting further trials within it.

Let the current minimum value of $z^*$ from (\ref{z}) be achieved at the left point of the $i^{\rm th}$ interval, i.e. $z^* = z_{i-1}$. 
As proven in \cite{Strongin2000}, according to the rule (\ref{R}) the following inequality will be true: 
\[
R(i) \geq \Delta_i \left( 1 - \frac{1}{r} \right)^2.
\]
Therefore, for the estimates of the characteristics $R_{loc}(i)$ and $R_{glob}(i)$ calculated with different parameters $r_{loc}$ and $r_{glob}$, the following relation will hold:
\[
	\Delta_i \left( 1 - \frac{1}{r_{glob}} \right)^2 > \Delta_i \left( 1 - \frac{1}{r_{loc}} \right)^2.
\]

Thus, when choosing the largest of the characteristics in accordance with
\[
R(i) = \max\{R_{loc}(i),R_{glob}(i)\}
\]
%as a rule,
the characteristic $R_{glob}(i)$ corresponding to the higher estimate of the Lipschitz constant will be chosen; the lower estimate (which speeds up the process of refining the current solution) will not be used.
However, if we multiply the lower estimate for the characteristic $R_{loc}(i)$ by the coefficient $\rho$ in accordance with (\ref{pho}), then such lower estimates will be equal, 
thus the choice of the characteristic $R_{loc}(i)$ corresponding to the lower estimate of the Lipschitz constant will become more likely.

@@

The advantages of the GSA-DL algorithm over its prototype are also confirmed by the operational characteristics of the algorithms as well. Assume a series of test problems to be solved. The results of solving the series can be presented by a function $p(k)$ featuring the fraction of the total number of problems solved in $k$ iterations. Such a function will be called the \textit{operational characteristic} of the algorithm.

@@

The lower curves in Fig. \ref{oper4} and Fig. \ref{oper5} feature the characteristics of the GSA method whereas the upper ones, those of the GSA-DL. Such relative positions of the curves show the algorithm with two estimates of the Lipschitz constant is much faster on average when solving the problem series than the algorithm using a single estimate of the constant.
%Russian
Note that to solve problems of all classes (except for the \textit{Hard} class at $N=5$) the GSA-DL method requires about half as many trials as the GSA. The deterioration of the results in the case of the \textit{Hard} class at $N=5$ is explained by the complexity of this class' functions, which have a large attraction region of local minima and a narrow attraction region of the global minimum. To correctly solve such problems, the GSA-DL method often uses a higher estimate of the Lipschitz constant, thus reducing the speed difference of the GSA and GSA-DL algorithms.
